---
title: "Set-up DESeq2 analysis - Answer Key"
authors: "Will Gammerdinger, Noor Sohail"
date: "Friday, September 5, 2025"
editor_options: 
  markdown: 
    wrap: 72
---


```{r}
#| label: load_data
#| echo: false
# Load libraries
library(Seurat)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(EnhancedVolcano)
library(RColorBrewer)
library(cowplot)
library(dplyr)
library(ggrepel)

seurat <- readRDS("data/BAT_GSE160585_final.rds")

meta_columns <- c("sample", "condition")
meta <- seurat@meta.data %>%
            select(meta_columns) %>%
            unique() %>%
            remove_rownames()

bulk <- AggregateExpression(
            seurat,
            return.seurat = TRUE,
            assays = "RNA",
            group.by = c("celltype", "sample", "condition")
)

n_cells <- seurat@meta.data %>% 
              dplyr::count(sample, celltype) %>% 
              rename("n"="n_cells")
n_cells$sample <- str_replace(n_cells$sample, "_", "-")

meta_bulk <- left_join(bulk@meta.data, n_cells)
rownames(meta_bulk) <- meta_bulk$orig.ident
bulk@meta.data <- meta_bulk

# Turn condition into a factor
bulk$condition <- factor(bulk$condition, levels=c("TN", "RT", "cold2", "cold7"))

bulk_vsm <- subset(bulk, subset= (celltype == "VSM") & 
                     (condition %in% c("TN", "cold7")))
```

# Exercise 1

Another cell type in this dataset that was particularly interesting to the authors were the **Pdgfr α+ adipose progentior cells (APCs)**.

1. Subset the `bulk` object to isolate only adipose progenitor cells for the TN and cold7 conditions. Assign it to variable called `bulk_APC`.

 _**Hint**: You may need to review `celltypes` to determine what this cell type is called in our data._ You can find unique celltypes with the code:
   
```{r}
#| label: unique_celltypes
# Find unique celltypes
celltypes <- sort(unique(seurat@meta.data[["celltype"]]))
```

::: callout-note
The abbreviations for the cell types can be found in the [project set-up lesson](02_setup_intro_dataset.qmd#celltype-annotation).
:::

```{r}
#| label: APC_subset
celltypes
# Compare TN vs cold7 in APC cells
bulk_APC <- subset(bulk, subset = (celltype == "AP")  & (condition %in% c("TN", "cold7")))
```

2. Plot the cell number distribution across samples. How do the numbers compare to VSM cells?

```{r}
#| label: APC_ncells_plot
# Visualize number of cells per condition
ggplot(bulk_APC@meta.data, aes(x = sample, y = n_cells, fill = condition)) +
  geom_bar(stat = "identity", color = "black") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Sample name", y = "Number of cells") +
  geom_text(aes(label = n_cells), vjust = -0.5)
```

::: {.callout-note collapse="true"}
# Click here for code to plot the two bar plots side-by-side

Note that this R code below uses the **ggpubr library**. In order to run this you will need to first install the package and then run:

```{r}
library(ggpubr)

# Plot VSM and APC side by side
plot_cell_number_vsm <- ggplot(bulk_vsm@meta.data, aes(x = sample, y = n_cells, fill = condition)) +
  geom_bar(stat = "identity", color = "black") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Sample name", y = "Number of cells") +
  geom_text(aes(label = n_cells), vjust = -0.5) +
  ggtitle("VSM") +
  theme(title = element_text(hjust = 0.5))

plot_cell_number_APC <- ggplot(bulk_APC@meta.data, aes(x = sample, y = n_cells, fill = condition)) +
  geom_bar(stat = "identity", color = "black") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Sample name", y = "Number of cells") +
  geom_text(aes(label = n_cells), vjust = -0.5) +
  ggtitle("APC") +
  theme(title = element_text(hjust = 0.5))

ggpubr::ggarrange(plot_cell_number_vsm, plot_cell_number_APC, nrow = 1,
                  common.legend = TRUE, legend = "right")
```

:::

Overall we see far fewer cells, by an order of magnitude (scale goes to 2,000 for VSM but only 600 for APC). There is also a different distribution: the counts for Sample-10 and Sample-9 go down relative to other samples, while Sample-8 goes up.


# Exercise 2

Create a DESeq2 object for the **Pdgfr α+ APCs** data as `dds_APC`.

```{r}
#| label: create_apc_dds_object
# Get count matrix
APC_counts <- FetchData(bulk_APC, layer="counts", vars=rownames(bulk_APC))

# Create DESeq2 object
# transpose it to get genes as rows
dds_APC <- DESeqDataSetFromMatrix(t(APC_counts),
                                colData = bulk_APC@meta.data,
                                design = ~ condition)

dds_APC
```