---
title: "Set-up DESeq2 analysis - Answer Key"
authors: "Will Gammerdinger, Noor Sohail"
date: "Friday, September 5, 2025"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_data_libraries
#| echo: false
# Load libraries
library(tidyverse)
library(clusterProfiler)
library(org.Mm.eg.db)
library(msigdbr)
library(DT)
dge_deseq2 <- read.csv("results/deseq2_VSM_cold7_vs_TN.csv")

# Untested genes have padj = NA, so let's keep genes with padj != NA
dge_deseq2_noNAs <- filter(dge_deseq2, padj != "NA" )

# Create background dataset for hypergeometric testing using all tested genes for significance in the results
all_genes <- as.character(dge_deseq2_noNAs$gene)

dge_vsm <- read.csv("results/findmarkers_vsm_cold7_vs_TN.csv")
```

# Exercise 1

Using the code above as a template, run the over-representation analysis on the significantly down-regulated genes from the pseudobulk analysis.

```{r}
#| label: ORA_sigDown
# Extract significant results for down-regulated
sigDown <- dplyr::filter(dge_deseq2_noNAs, padj < 0.05, log2FoldChange < 0)
sigDown_genes <- as.character(sigDown$gene)

# Run GO enrichment analysis 
egoDown <- enrichGO(gene = sigDown_genes, 
                    universe = all_genes,
                    keyType = "SYMBOL",
                    OrgDb = org.Mm.eg.db, 
                    ont = "BP", 
                    pAdjustMethod = "BH", 
                    qvalueCutoff = 0.05, 
                    readable = TRUE)

# Output results from GO analysis to a table
cluster_summaryDown <- data.frame(egoDown)

# Save results
write.csv(cluster_summaryDown, "results/clusterProfiler_VSM_TNvsCold7_downregulated.csv")
```

   * How many significant terms do you find?

```{r}
#| label: sigDown_terms
# Number of significantlt down-regulated terms
nrow(cluster_summaryDown)
```

   * What are some of the prominent biological processes that are observed?
   
```{r}
#| label: view_prominent_BPs
head(cluster_summaryDown$Description)
```
   
# Exercise 2

Now that we have run through functional analysis with the results from Pseudobulk DE, let's see what results we derive from the DGE lists from our [FindMarkers DE analysis](03_DEanalysis_using_FindMarkers.qmd).

Create a significant DE genes data frame from the FindMarkers results with an added fold change criteria to reduce the gene list size. You can do this by running the code below:

```{r}
#| label: create_sig_DE_findmarkers
# Create significant DE genes from FindMarkers()
sig_fc_dge <- dge_vsm %>% dplyr::filter(p_val_adj < 0.05, abs(avg_log2FC) > 1)
```

1. Use this gene list to run an over-representation analysis. Be sure to separate genes into up- and down-regulated first. Also keep in mind that the background gene dataset is different than for the DESeq2 analysis. 

```{r}
#| label: run_ORA
# Create background dataset for hypergeometric testing using all tested genes for significance in the results
all_genes_fm <- dge_vsm$gene

# Extract significant results for up- and down-regulated
sigUp_fm <- dplyr::filter(sig_fc_dge, avg_log2FC > 0)
sigUp_fm_genes <- sigUp_fm$gene
sigDown_fm <- dplyr::filter(sig_fc_dge, avg_log2FC < 0)
sigDown_fm_genes <- sigDown_fm$gene

# Run GO enrichment analysis 
egoUp_fm <- enrichGO(gene = sigUp_fm_genes, 
                     universe = all_genes_fm,
                     keyType = "SYMBOL",
                     OrgDb = org.Mm.eg.db, 
                     ont = "BP", 
                     pAdjustMethod = "BH", 
                     qvalueCutoff = 0.05, 
                     readable = TRUE)

egoDown_fm <- enrichGO(gene = sigDown_fm_genes, 
                       universe = all_genes_fm,
                       keyType = "SYMBOL",
                       OrgDb = org.Mm.eg.db, 
                       ont = "BP", 
                       pAdjustMethod = "BH", 
                       qvalueCutoff = 0.05, 
                       readable = TRUE)

# Output results from GO analysis to a table
cluster_summaryUp_fm <- data.frame(egoUp_fm)
cluster_summaryDown_fm <- data.frame(egoDown_fm)
```

What are the top terms enriched among up-regulated genes? 

```{r}
#| label: up_terms
# Print top up-regulated terms from FindMarkers()
head(cluster_summaryUp_fm$Description)
```

What are the top terms enriched among down-regulated genes?

```{r}
#| label: down_terms
# Print top down-regulated terms from FindMarkers()
head(cluster_summaryDown_fm$Description)
```

2. How do these results compare with what we observed from the Pseudobulk DE functional analysis?

In the pseudobulk results, we also saw terms related to the extracellular matrix and cell adhesion upregulated. However, the downregulated results are different: for FindMarkers we see muscle and fat development, whereas for pseudobulk we saw RNA splicing and translation downregulated.